---
title: "Covid-19 in England by area type"
author: "Matt Kerlogue"
date: "2020-10-26"
output:
  html_document:
    code_folding: hide
    dev: svg
    fig_width: 10
    fig_height: 8
    fig_caption: true
    highlight: zenburn
    toc: true
    toc_float: true
---

**This RMarkdown notebook looks at the prevalence of Covid-19 in England by area type. It is exploratory and developed out of personal curiosity, it does not reflect official government policy or the views of my employer.**

There has been speculation about whether the recent rise in Covid-19 cases in England has been driven by students. This has come about through cursory analysis of the case-rates for MSOAs that have the term "university" in their 'colloquial' name (as defined by the House of Commons Library's MSOA names reference). I thought that there must be a more data driven approach to identifying "student" MSOAs than simply relying on their names.

```{r setup, echo=FALSE, include=FALSE}
knitr::opts_chunk$set(
  dev = "svg",
  out.width = "100%"
)

library(tidyverse)
msoa_data <- read_csv("data/msoa_data.csv", col_types = cols(.default = col_character()))
msoa_data <- msoa_data %>% mutate(all_ages = as.numeric(all_ages),
                                  across(starts_with("wk"), as.numeric))

msoa_cases_by_type <- msoa_data %>%
  group_by(subgroup_code, subgroup_name) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE), .groups = "drop")

subgroups <- msoa_cases_by_type$subgroup_code %>% set_names(msoa_cases_by_type$subgroup_name)

msoa_cases_by_type <- msoa_cases_by_type %>%
  mutate(subgroup = factor(subgroup_code, levels = subgroups, labels = str_wrap(names(subgroups), 15)))

type_pop <- msoa_cases_by_type %>%
  select(subgroup, all_ages)

```

## University cases

Let's use the MSOA names 'method' first of all to see if there's value in it. We are currently in week 43, so week 42 is the latest 'full' data for MSOAs. As this is something that people have been talking about for a couple of weeks, let's go back and look at the data for week 40 (the week to `r lubridate::ymd("2020-01-02") + lubridate::weeks(40)`). If we look at the ten MSOAs with the highest case rate (cases per 100,000 people) then four of these include the word "university" in their name.

```{r}

university_msoa <- msoa_data %>%
  mutate(university_area = str_detect(tolower(msoa11_hclnm), "university"))

university_cases <- university_msoa %>%
  select(msoa11_cd, msoa11_hclnm, all_ages, wk_40) %>%
  mutate(rate = wk_40/(all_ages/100000))
  

university_cases %>% 
  slice_max(order_by = rate, n = 10) %>%
  arrange(-rate) %>%
  mutate(msoa11_hclnm = if_else(
    str_detect(tolower(msoa11_hclnm), "university"),
    paste0("<strong>", msoa11_hclnm, "</strong>"), 
    msoa11_hclnm)) %>%
  mutate(
    all_ages = formattable::comma(all_ages, 0),
    rate = formattable::comma(all_ages, 2)
  ) %>%
  knitr::kable(escape = FALSE, format = "html") %>%
  kableExtra::kable_styling()
    

```

If we look at all MSOAs over time by whether they have "university" in their name or not we can see that there has been a clear spike in the case rate for "university" MSOAs.

```{r, fig.height=5}

university_cases_graph_dt <- msoa_data %>%
  mutate(university = factor(if_else(str_detect(tolower(msoa11_hclnm), "university"),
                             "University MSOA", "Other MSOA"), 
                             levels = c("University MSOA", "Other MSOA"))) %>%
  group_by(university) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE), .groups = "drop") %>%
  pivot_longer(cols = c(starts_with("wk")), 
               names_to = "week", values_to = "cases") %>%
  mutate(week = as.numeric(str_remove(week, "wk_")),
         rate = cases/(all_ages/100000)) %>%
  filter(week > 26)

ggplot(university_cases_graph_dt, aes(x = week, y = rate, colour = university)) +
  geom_line() +
  scale_colour_manual(values = c("#d7191c", "#2c7bb6")) +
  labs(x = "Week", y = "Covid-19 case rate (per 100,000)") +
  mattR::theme_lpsdgeog() +
  theme(legend.title = element_blank(),
        legend.position = "top")

```

However there are only `r  sum(grepl("university",tolower(msoa_data$msoa11_hclnm)))` MSOAs with the word "university" in their name out of `r formattable::comma(nrow(msoa_data),0)` MSOAs, with a population of `r formattable::comma(sum(msoa_data[grepl("university",tolower(msoa_data$msoa11_hclnm)), "all_ages"]), 0)` people, a very small fraction of the data.

## Using Area Classifications

There are a number of geo-demographic classification systems that cluster local areas based on a range of key characteristics. The most famous (Mosaic and Acorn), but there is an open classification developed by the ONS and University College London from the 2011 Census: the Output Area Classification. Five of the 69 sub-groups in the classification specifically reference students in their names.

```{r}
msoa_data %>%
  distinct(subgroup_code, subgroup_name) %>%
  filter(str_detect(tolower(subgroup_name), "student")) %>%
  arrange(subgroup_code) %>%
  knitr::kable(format = "html") %>%
  kableExtra::kable_styling()
```

The 2011 Census was almost a decade ago and its successor is due to take place in less than six months time. However, the last Census remains the best available public data on the geographical distribution of students. While there is no doubt going to have been some changes, given that universities are relatively fixed geographical entities it is likely that the general pattern arising from the 2011 Census is likely to be good enough for us here.

### Aggregating the area classifications

The 2011 Output Area Classifications (OAC) are based on the smallest unit of Census geography, the Output Area. There is an aggregation of these classifications for local authorities but not for intermediate geographies such as MSOAs. MSOAs have been assigned to one of the 69 area OAC sub-groups using the following method:

-   Obtain the 2019 Mid-Year Population Estimates for Output Areas in England
-   Join the population estimates to the OAC data
-   Link Output Areas to their containing MSOA
-   Calculate the proportion of the MSOA population that is from each subgroup (`subgroup_pop_pc`)
-   Calculate the proportion of Output Areas within an MSOA allocated to each subgroup (`subgroup_oa_pc`)
-   Create a score by multiplying these two proportions (`scores = subgroup_pop_pc * subgroup_oa_pc`)
-   Assign the MOSA to the subgroup with the highest score

```{r}

msoa_data %>%
  select(subgroup_code, subgroup_name, all_ages) %>%
  add_count(subgroup_code, subgroup_name, name = "n_msoas") %>%
  add_count(subgroup_code, subgroup_name, wt = all_ages, name = "population") %>%
  select(-all_ages) %>%
  distinct() %>%
  filter(str_detect(tolower(subgroup_name), "student")) %>%
  mutate(population = formattable::comma(population, digits = 0)) %>%
  arrange(subgroup_code) %>%
  knitr::kable(format = "html") %>%
  kableExtra::kable_styling()

```

This gives us `r  sum(grepl("^student ",tolower(msoa_data$subgroup_name)))` MSOAs that are likely to be largely student populations (Student Communal Living and Student Digs) with a population of `r formattable::comma(sum(msoa_data[grepl("^student ",tolower(msoa_data$subgroup_name)), "all_ages"]), 0)`, and a further `r sum(grepl("student",tolower(msoa_data$subgroup_name))) - sum(grepl("^student ",tolower(msoa_data$subgroup_name)))` areas that are likely to have significant student populations with a population of `r formattable::comma(sum(msoa_data[grepl("student",tolower(msoa_data$subgroup_name)), "all_ages"]) - sum(msoa_data[grepl("^student ",tolower(msoa_data$subgroup_name)), "all_ages"]), 0)`. A more substantial set of areas to base our analysis on than just relying on the MSOA name.

Another potential arguement in favour is that there are 11 MSOAs which contain the word "university" in their name but were not assigned to a student OAC, perhaps a university institution is based within that MSOAs but it might not be where students are actually living.

``` {r}
msoa_data %>% 
  filter(str_detect(tolower(msoa11_hclnm), "university")) %>% 
  filter(str_detect(tolower(subgroup_name), "student", negate = TRUE)) %>% 
  select(msoa11_cd, msoa11_hclnm, subgroup_code, subgroup_name) %>%
  knitr::kable(escape = FALSE, format = "html") %>%
  kableExtra::kable_styling()
```


### Cases by area classification

First off, looking at raw case numbers across all sub-group types, we can see that there are some spikes in the sub-groups related to students, but also in other area types too (White Suburban Communities and Semi-Detached Ageing).

```{r}
type_cases_graph_dt <- msoa_cases_by_type %>%
  select(subgroup, starts_with("wk")) %>%
  pivot_longer(cols = -subgroup, names_to = "week", values_to = "cases") %>%
  mutate(week = as.numeric(str_remove(week, "wk_"))) %>%
  filter(week > 26)

ggplot(type_cases_graph_dt, aes(x = week, y = cases, colour = subgroup)) +
  geom_line() +
  facet_wrap(~subgroup) +
  scale_y_continuous(limits = c(0,9000),
                     breaks = c(0, 4500, 9000),
                     labels = c("0", "4.5k", "9k")) +
  labs(x = "Week", y = "Number of cases") +
  mattR::theme_lpsdgeog() +
  theme(legend.position = "none",
        axis.text.y = element_text(hjust = 1),
        panel.grid.minor = element_blank(),
        strip.text = element_text(size = 7))

```


This first chart uses a fixed y-axis across each area, so we might be missing patterns in areas with smaller populations (and thus smaller number of cases), letting the y-axis vary for each group. This shows us that the rise in cases is happening across all area types.

```{r}
ggplot(type_cases_graph_dt, aes(x = week, y = cases, colour = subgroup)) +
  geom_line() +
  facet_wrap(~subgroup, scales = "free_y") +
  labs(x = "Week", y = "Number of cases") +
  mattR::theme_lpsdgeog() +
  theme(legend.position = "none",
        panel.grid.minor = element_blank(),
        panel.grid.major.y = element_blank(),
        axis.text.y = element_blank(),
        panel.border = element_rect(colour = "grey80", fill = NA),
        strip.text = element_text(size = 7))
```

However, the best way of comparing area is to look at the case rate (per 100,000). As with the 'university' example this shows clear spikes for the "Student Communal Living" and "Student Digs".

```{r}

type_rate_graph_dt <- type_cases_graph_dt %>%
  left_join(type_pop, by = "subgroup") %>%
  mutate(rate = cases / (all_ages/100000)) %>%
  filter(week > 26)

ggplot(type_rate_graph_dt, aes(x = week, y = rate, colour = subgroup)) +
  geom_line() +
  scale_y_continuous(breaks = c(0, 1000), labels = c("", "1k")) +
  labs(x = "Week", y = "Covid-19 case rate (per 100,000)") +
  facet_wrap(~subgroup) +
  mattR::theme_lpsdgeog() +
  theme(legend.position = "none",
        panel.border = element_rect(colour = "grey80", fill = NA),
        panel.grid.minor = element_blank(),
        strip.text = element_text(size = 7),
        axis.text.y = element_text(hjust = 1))

```

To more easily visualise this, let's compare area types in a single graph, let's combine the "Student Communal Living" and "Student Digs" area types into a "Mainly students" neighbourhood category, the three other subgroups with students in their name into a "Students and others" category, and then all other neighbourhoods.

```{r, fig.height=5}

student_neighbourhoods <- msoa_cases_by_type %>%
   mutate(neighbourhood = case_when(
     subgroup_code == "2a1" ~ "Mainly students",
     subgroup_code == "2a2" ~ "Mainly students",
     str_detect(subgroup_name, "student|Student") ~ "Students and others",
     TRUE ~ "Other neighbourhoods"),
     neighbourhood = factor(neighbourhood, 
                            levels = c("Mainly students", 
                                       "Students and others", 
                                       "Other neighbourhoods"))) %>%
  select(neighbourhood, all_ages, starts_with("wk")) %>%
  group_by(neighbourhood) %>%
  summarise(across(where(is.numeric), sum, na.rm = TRUE), .groups = "drop")

student_rates <- student_neighbourhoods %>%
  select(neighbourhood, starts_with("wk_")) %>%
  pivot_longer(cols = -neighbourhood, names_to = "week", values_to = "cases") %>%
  mutate(week = as.numeric(str_remove(week, "wk_"))) %>%
  left_join(select(student_neighbourhoods, neighbourhood, all_ages), 
            by = "neighbourhood") %>%
  mutate(rate = cases / (all_ages/100000)) %>%
  filter(week > 26)

ggplot(student_rates, aes(x = week, y = rate, colour = neighbourhood)) +
  geom_line() +
  scale_x_continuous(minor_breaks = 26:42) +
  scale_colour_manual(values = c(
    "Mainly students" = "#d7191c",
    "Students and others" = "#fdae61",
    "Other neighbourhoods" = "#2c7bb6"
  )) +
  labs(x = "Week", y = "Covid-19 case rate (per 100,000)") +
  mattR::theme_lpsdgeog() +
  theme(panel.border = element_rect(colour = "grey80", fill = NA),
        legend.position = "top",
        legend.title = element_blank())

```

The case rates in the student areas are notably higher than in other area types, and happened before the rise in other area types.

### Potential applications
As mentioned at the top of this report, this analysis is purely due to my own personal curiosity in this issue. However, it is possible that using the Output Area Classification with MSOA case data could enable more targeted action to be taken, the chart above clearly shows us that the case rate in what are the OAC suggests are predominately student areas was starting in week 36, and certainly by week 38. A subsequent rise then started in other student areas and at the time of writing appears to be extending to other area types.

The following chart shows the case rates for weeks 36-42 by area type but with the y-axis limited to a rate of 0-200 per 100,000 per group, let's also add a dashed horizontal line to indicate a rate of 100 cases per 100,000 and a thick vertical line to denote week 40.

```{r}

ggplot(filter(type_rate_graph_dt, week >= 35), aes(x = week, y = rate, colour = subgroup, fill = subgroup)) +
  geom_line() +
  geom_area(alpha = 0.6) +
  geom_hline(yintercept = 100, linetype = "dashed", size = 0.2, colour = "grey40") +
  geom_vline(xintercept = 40, size = 0.8, colour = "grey40") +
  labs(x = "Week", y = "Covid-19 case rate (per 100,000)") +
  scale_y_continuous(limits = c(0, 200), oob = scales::oob_keep) +
  scale_x_continuous(minor_breaks = 35:42) +
  facet_wrap(~subgroup, scales = "free_y") +
  mattR::theme_lpsdgeog() +
  theme(legend.position = "none",
        panel.border = element_rect(colour = "grey80", fill = NA),
        panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank(),
        strip.text = element_text(size = 7),
        axis.text.y = element_blank())

```

There are still some area types where the case rate is below 100, but not many and the majority of places appear to be on a trajectory towards breaching that rate. We can also see that the student neighbourhoods rose above the case rate of 100 per 100,000 earlier than week 40, while in other area types this clearly happened at or after week 40.

There are however some limitations with the available data that would either need rectifying/caveating if this were to be used more robustly in decision making:

* Local geography is going to be more important in addressing the spread, containment and response to the Covid-19 pandemic. Analysis may only be useful in post-event analysis rather than as an in-the-moment tool.
* The application of the OAC to MSOAs should be validated, the current assignment method is a quick and dirty approximation. Ideally, an MSOA specific classification would be developed that ensures a more appropriate assignment of MSOAs either to the existing classification or a one-off bespoke classification.
* Creating a bespoke MSOA classification for the purposes of responding to Covid-19 might have its advantages, for example by combining other data being used in understanding the pandemic such as travel to work/inter-area mobility, physical infrastructure (e.g. communal establishments, hospitals) and other measures.
* Data from the ONS Annual Population Survey could provide be a source of data to either validate/update information about the presence of students - there is an option within Nomis to extract some data by the 2011 OAC, but this did not allow for analysis of the numbers in higher eduction by OAC area type.


## About the data

The data used in this report comes from the Office for National Statistics (ONS), Public Health England (PHE). PHE data also includes data from the House of Commons Library (HOCL). ONS and PHE data are Crown Copyright and used under the terms of the [Open Government Licence](https://www.nationalarchives.gov.uk/doc/open-government-licence/version/3/), HOCL data are Parliamentary Copyright and used under the terms of the [Open Parliament Licence](https://www.parliament.uk/site-information/copyright-parliament/open-parliament-licence/).

The data used in this report are:

* ONS 2019 Mid-Year Population Estimates for Mid Layer Super Output Areas (MSOAs) and Output Areas [link](https://www.ons.gov.uk/peoplepopulationandcommunity/populationandmigration/populationestimates/datalist?filter=datasets)
* ONS 2011 Census Output Area Classification [link](https://www.ons.gov.uk/methodology/geography/geographicalproducts/areaclassifications/2011areaclassifications/datasets)
* PHE Covid-19 case data for MSOAs [link](https://coronavirus.data.gov.uk/about-data#cases-by-middle-super-output-area-msoa)
